From Questions to Queries – Enhancing the Curt System by a Theory and Implementation of Embedded WH-Questions

### 2008-08-25 thesis/paper.tex
Cleaned up section 1 and 2. introduced citations

\newcommand{\abbr}{\textsf} % how to render abbreviations like IL or Ty2
\newcommand{\stress}{\textbf} % put stress somewhere (as opposed to \emph)
\newcommand{\term}{\textsf} % introduce a term
\newcommand{\code}{\texttt} % layout of in-line code snippets
\newcommand{\pn}{\textsf} % how to render proper names like Shrdlu
\newcommand{\url}[1]{\code{http://#1}} % how urls appear
\newcommand{\example}{\textit} % in-text examples

@@

%theorem environments
\theoremstyle{remark} \newtheorem*{termin}{Definition} % terminology. Could use a
%better title fixme

@@

\section*{Introduction}

Growing efforts have thus been made to provide a technology that is (going to
be) able to extract logical formulae from sentences of ordinary language and
establish relations between objects of its logical representation. In addition,
providing a means to compute with said representations is will eventually enable
the machine-processing of meaning.  In doing so, the ultimate goal may be to
enable a man-made system to \emph{understand} what a humans communicate.

@@

Development of exhaustive systems in this area appears to be a daunting task of
quite some scope. Therefore most approaches have been centered on providing
coverage of at least a fragment of natural language. Many such systems exist and
they prove to be increasingly successful. Some of them will be discussed here
and presenting an extension to one system in particular is the main matter of
this Bachelor's thesis.

In doing so, the present work will aim to provide the \curt system of
\cite{blackburnbos:cl1} with the mechanics of dealing with more than one
partition of its logical domain, viz. the ability to handle more than one
\term{possible} or \term{accessible world}. As a ground for development of this
work, the treatment of \term{embedded questions} was chosen as one that
naturally demands a possibility to account for more than one division in a
system's \emph{understanding} or representation of a universe of objects and
relations.

As

The first section gives a conceptual overview, introducing the \curt
system and comparing it to other similar systems in existence. It then goes on
to explain the target domain of natural language the implementation has to
account for.
The next section concerns itself with a more detailed analysis of the theoretic
assumptions of this implementation and justifies some choices and restrictions
that had to be made because of the limited scope of this analysis. While the
third section will display and clarify the actual implementation on top of the
\curt system, the last section will discuss the results of this work and also
give a prospect on possible further work on the subject.

\subsection*{Acknowledgements}

Frank... who else?

@@

The basis of

%fixme: to be formalized
Where we use ``natural language'', usually the English language will be treated
in practice, though it would be nice to come up with treatments of other
languages. We abstract away from pragmatic affairs and concern ourselves with
mostly the purely semantic/logic domain, so this should be feasible to achieve.

A brief overview of the premisses for the theory (which itself lays ground to
the implementation).

@@

\pn{Curt}, a system written in \prol that is able to interpret basic utterances
in natural language and store them in expressions of first order logic, is
presented in \cite[chapter 6]{blackburnbos:cl1}. While relatively limited when
compared to other systems presented in \ref{sec:comparision} in terms of lexical
and grammatical coverage, it provides a sound basis for extended work because of
its coherent design and relative simplicity.

Besides parsing and translating utterances into logic, \curt is able to perform
basic checks for \emph{consistency} and \emph{informativity} on the resulting
formulae -- within the bounds of the tools it uses.

% FIXME: don't really like this footnote
\footnotetext{Here, the notion of \emph{consistency} is meant as \emph{validity}
of the input and \emph{informativity} denotes \emph{validity}. Thus, if
\Disc, the discourse so far and $\phi$ a new input token (fully parsed
sentence), then consistency means $\Disc \to \phi$ holds and informativity
means $\Disc \nvDash \phi$.}

It can thus detect incoherent discourse and would reject a sequence of
sentences such as \emph{``Mary likes every man. Peter is a man. Mary does not
like Peter''}, claiming it found a contradiction in its input. For every input
token, \pn{Curt} will also try to infer whether the current discourse would
entail the information given in its newest token and is thus irrelevant with
regard to previous utterances – effectively allowing it to
reduce the size of its internal representation of the discourse by not adding
clauses to it which would not contribute to the overall informational value
already present in the program.

The system is introduced during the final chapter of \cite{blackburnbos:cl1},
incrementally increasing in features and complexity. Support for external
inferencing via theorem provers and model builders is added in a step-by-step
fashion and finally an ontology and a straight-forward way of parsing and
answering direct questions for noun phrases conclude its development. The theory
and implementation of this paper is meant as a replacement and enhancement for
the latter feature. This will also happen in an incremental manner and new
features will be introduced one at a time.

\subsection{General Comparison of the Curt System}\label{sec:comparision}

This section will briefly discuss how the features of the Curt system compare to
other similar systems or systems with the same goal, namely implementing a
system that is able to model a logic representation of human language.

\subsubsection{Early Attempts: \pn{Shrdlu} and \pn{Cyc}}
%TODO: need some reference for this section!

Interest in deep semantic parsing of natural language has existed since the very
beginning of the informational age. One of the first systems to achieve public
recognition, the \pn{Shrdlu} system

While the \pn{Curt} system is unique in its approach to deep semantic analysis, its
architecture has been crafted in a pragmatic manner. The authors chose to use
external tools where it was possible in order to avoid reimplementation of
existing functionality and be able to focus on novel parts of the
implementation\footnote{The name of the \pn{Curt} system reflects this choice.
It stands for \emph{\textbf{C}lever \textbf{u}se of \textbf{r}easoning
\textbf{t}ools}.}. This also allows for great flexibility in adopting new
solutions and improvements made in the field of inferencing tools for first
order predicate logic.

@@

At the end of \cite{blackburnbos:cl1}, \curt is able to cope with a small
grammar and provides some features for interacting with humans in natural
language. It does only understand basic extensional transitive verbs - modal
verbs or the treatment of questions are not implemented. Anaphora resolution
still remains an open issue. The grammar is rather limited and \curt only comes
with a basic vocabulary oriented around the setting of a popular movie.

The aim of this paper is to introduce the reader into question semantics as it
will be relevant for implementing a basic capability of treating intensional
verbs like `to \stress{know} whether something is the case' and `to
\stress{believe} that something might be the case' and then present such an
implementation within the framework of the \curt system. Further and more
advanced investigations will be dealt with at the end of this discussion or left
for further research.

\section{Theory}

@@

Following \cite{gs:q}, we will not use the natural natural language word
``question'' but distinguish between \term{questions},
\term{interrogative sentences} and \term{interrogative acts} in the following
way:

\begin{termin}
  An \term{interrogative} is a certain type of sentence in natural language
  distinguished by certain features\footnote{To quote \cite{gs:q}: \quote{[An
  interrogative sentence is] characterized by word order, intonation, question
  mark, the occurrence of interrogative pronouns}.}
  whereas an \term{interrogative act} refers to the utterance of an
  \emph{interrogative}. Finally, a 
  \term{question} will denote the semantic entity uttered by the
  \emph{interrogative act}, whichever form it may have.
\end{termin}

\abbr{Ty2} is two-sorted type logic\footnote{See \cite{gallin:ty2} for a
discussion and \cite{gs:sqpa} for its usage in describing the semantics of
questions}. \abbr{IL} is intensional logic\footnote{As used by \cite{ptq}}. 
An embedded \wh-phrase looks
like:%fixme

\ex. John \stress{knows} \emph{whether} good examples are hard to come up with.

After having elaborated on the grounds of the implementation, we will now
display the implementation's underlying concepts. % FIXME this sentence is too short

%OUTLINE Presenting a conceptual overview, while arguing for and against stuff.

@@

\subsection{Existing Implementations of Similar Systems}\label{sec:othercrap}

Enhancing \curt by an implementation dealing with questions in natural language
sparks interest in whether there have been similar efforts made in the past. No
further extensions to the \curt system aiming at comparable goals are known to the
author at the time of writing. However, there is a lively interest in providing
so-called question answering or QA-systems since the early stages of computer
science. % fixme; cite some ancient 60's paper
This section discusses a few of the more elaborate approaches: first an early
but already quite impressive implementation is shown, followed by a newer approach
using embodied agents and finally an overview over the many database-based
QA-systems is given. Finally, all this is compared what the current extension to
\curt can do.

\subsubsection{Godot, the Talking Robot}

\subsubsection{CHAT-80}

\subsubsection{Modern Intensional Database Approaches}

@@

\prol programs can be self-modifying, or dynamic. This means that the program is
able to change its own state and thus store information.

@@

\curt parses sentences of natural language using an enhanced version of \pn{Cooper
storage}\footnote{See \cite{cooper:storage2} for the original implementation.} as
presented in \cite{keller:storage}. \cite{blackburnbos:cl1} call this version
\pn{Keller storage}. That name will be used from here % FIXME sucks. Please change
to refer to this particular implementation as presented in
\cite{blackburnbos:cl1}.

\pn{Keller storage} is already quite capable
of dealing with more complex constituents such as nested noun phrases, but it
has its limitations. In particular, it can only handle quantifier scope
ambiguities and does not take negation into account. Moreover, it overgenerates
with some quantifier scope ambiguities\footnote{\cite{blackburnbos:cl1} mention
sentences such as \ref{ex:kellersucks}, where Keller storage produces six
instead of the desired five readings.
\ex. One criminal knows every owner of a hash bar\label{ex:kellersucks}

Clearly, \example{a hash bar} cannot take scope over \example{every owner}, but
Keller storage predicts this reading.}.

In order to cope with more complicated quantifier configurations, where
constraints need to be enforced, and with other kinds of ambiguities,
Blackburn and Bos present an approach utilizing underspecified representations and
dominance relations, \pn{Hole Semantics}\footnote{Also presented in % todo:cite
%Bos' paper
}. Since, however, none of the cases discussed so far would require the enhanced
expressive power of \pn{Hole Semantics} it was decided to go on using \pn{Keller
Storage} instead, since this is what the original implementation of the \curt
system relies on.
%todo: fix capitalization in Keller storage all over the document

@@

\section{Discussion}

We have seen, the Curt system is quite capable of treating questions this way.
Having implemented modality and possible worlds in such a fashion, let's go on
and implement some even cooler stuff (tense, aspect, negation, comparatives,
anaphora). Basing all this on a nice implementation of DRT (discussed in
\cite{kampreyle:drt}) like the one given in \cite{blackburnbos:cl2} might be
interesting.

>>>>

\newcommand{\abbr}{\textsf} % how to render abbreviations like IL or Ty2
\newcommand{\stress}{\textbf} % put stress somewhere (as opposed to \emph)
\newcommand{\term}[1]{\textsf{\textbf{#1}}} % introduce a term
\newcommand{\code}{\texttt} % layout of in-line code snippets
\newcommand{\pn}{\textsf} % how to render proper names like Shrdlu
\newcommand{\url}[1]{\code{http://#1}} % how urls appear
\newcommand{\example}{\textit} % in-text examples

@@

%theorem environments
\theoremstyle{remark} \newtheorem*{termin}{Terminology} % terminology.

@@

\section{Introduction}

Growing efforts have thus been made to provide a technology that is (going to
be) able to extract logical formulae from sentences of ordinary language and
establish relations between objects of its logical representation. In addition,
providing a means to compute with said representations will eventually enable
the machine-processing of an abstract concept of meaning. In doing so, the
ultimate goal may be to enable a man-made system to \emph{understand} what a
humans communicate.

@@

Development of exhaustive systems in this area appears to be a daunting task of
quite some scope. Therefore most approaches have been focused on providing
coverage of a small fragment of natural language often with respect to a
particular domain. Many such systems exist and they prove to be increasingly
successful. Some of them will be discussed here and presenting an extension to
one system in particular is the main matter of this Bachelor's thesis.

\subsection{Premises}

The present work will aim to provide the \curt system of
\cite{blackburnbos:cl1} with the mechanics of dealing with more than one
partition of its logical domain, viz. the ability to handle more than one
\term{possible} or \term{accessible world}. As a ground for development of this
work, the treatment of \term{embedded questions} was chosen as one that
naturally demands a possibility to account for more than one division in a
system's \emph{understanding} or representation of a universe of objects and
relations.

As the scope of a Bachelor's thesis is limited by its nature, not all of the
theoretical background can be understood without background knowledge in some
rather divergent domains. In particular, a basic familiarity with traditional
\term{natural language semantics} and \term{extensional} as well as
\term{intensional logic} frameworks will be assumed. Furthermore, a basic
acquaintance with \term{eroteric logic} will be helpful in understanding some of
the reasoning in section \ref{sec:theory}.

Some knowledge of the \prol programming language and a good overview of the
architecture of the \curt system is imperative in order to understand the
implementation itself.

The first section gives a conceptual overview, introducing the \curt
system and comparing it to other similar systems in existence. It then goes on
to explain the target domain of natural language the implementation has to
account for.
The next section concerns itself with a more detailed analysis of the theoretic
assumptions of this implementation and justifies some choices and restrictions
that had to be made due to the limited scope of this analysis. While the
third section will display and document the actual implementation on top of the
\curt system, the last section will discuss the results of this work and also
give a prospect on possible further work on the subject.

@@

todo: write something here... :-(

@@

\curt, a system written in \prol that is able to interpret basic utterances
of English and store them in expressions of first order logic, is
presented in \cite[chapter 6]{blackburnbos:cl1}. While relatively limited when
compared to other systems presented in \ref{sec:comparison} in terms of lexical
and grammatical coverage, it provides a sound basis for extended work because of
its coherent design and relative simplicity.

Besides parsing and translating utterances into logic, \curt is able to perform
basic checks for \emph{consistency} and \emph{informativity} on the resulting
formulae -- within the bounds of the tools it uses.
It can thus detect incoherent discourse and would reject a sequence of
sentences such as \emph{``Mary likes every man. Peter is a man. Mary does not
like Peter''}, claiming it found a contradiction in its input. The discourse is
then said to be \term{inconsistent} with the new input.

For every input
sentence, \pn{Curt} will also try to infer whether the current discourse would
entail the information it contains and would thus be \term{uninformative} with
respect to previous utterances – effectively allowing it to
reduce the size of its internal representation of the discourse by avoiding to
add clauses to its database that would not contribute to the overall informational value
already present in the program.

The final chapter of \cite{blackburnbos:cl1} introduces the implementation
incrementally, presenting new versions of \curt that add to the software's
feature set. Support for external inferencing via theorem provers and model
builders is added right after the first naïve implementations and
finally an ontology and a straight-forward way of parsing and answering direct
questions for noun phrases conclude its development. The theory and
implementation of this paper is meant as a replacement and enhancement for the
latter feature. This will also happen in an incremental manner and new features
will be introduced one at a time.

\subsection{Comparison of the Curt System}\label{sec:comparison}

Generally, question answering and semantic parsing are not related and seldom
interdependent in current implementations of similar systems, with a few notable
exceptions.

\cite{chat} are among  the first to attempt creating a question answering system
with deep semantic coverage. Their effort has been focused on the geographical
domain, and their implementation, dubbed \pn{CHAT-80}, is written entirely in
\prol. The lack of hardware resources of that time forced the system to be
written with efficiency in mind. Moreover, it had to maintain a statical
database of facts which was hand-crafted. The \curt system is built on  an
entirely different set of premises and aims to provide a proof-of-concept of a
more general purpose system that is also able to extend its knowledge
dynamically.

Soon after \pn{CHAT-80}, two entirely different types of systems emerged. The
first type is probably the most promising to be used on a wider scale, given
today's technological standards. Implementations are usually based on
\term{information retrieval} and the generation of question/answer pairs. Such
systems are generally referred to as a \abbr{QA} (\term{Question Answering})
systems.  Examples implementations of such systems are discussed in \cite{rama},
\cite{iesqa} and many others. Most of these seem to provide a good lexical and
grammatical coverage due to their large input corpora and
shallow interpretation techniques. The latter is also the \abbr{QA} technique's
greatest shortcoming, as no deep analysis is  performed and usually
only factoids of less then a couple of words in length can be supplied as
answers.\footnote{But see \cite{aqabtf} for an approach that aims to cover a
wider range of possible answers.}

In contrast, \pn{CHAT-80} runs all its queries against a pre-built database
without the ability to acquire new knowledge from the discourse. While a
\abbr{QA} implementation will typically just acquire knowledge by consuming
more input and analysing its statistical properties, programs employing deep
semantic analysis will have to check their input for logical contradictions in
to maintain their database in a reasonable\footnote{FIXME pun intended, but
appropriate? %todo
} state.

\cite{prontoqa} introduce the \pn{Pronto QA System}, which couples traditional
corpus-based \abbr{QA} techniques with deep semantic parsing based on
\term{Discourse Representation Theory}.

A very interesting project, called \pn{Godot} is introduced in \cite{godot}.
\pn{Godot} is a so-called \term{embodied agent}, a robot with a sophisticated
software driving interaction with the outer world. As such its conception of
discourse not only relies on spoken dialogue, but also on sensory input.
\pn{Godot} is focused on executing orders -- after evaluating their feasibility
in its particular discourse model -- and does so far not support the answering
of questions about its state.

@@

At the end of \cite{blackburnbos:cl1}, \curt is able to cope with a small
grammar and provides some features for interaction in natural
language. It does only understand basic extensional transitive and intransitive
verbs and its support for question answering remains limited.
Anaphora resolution still remains an open issue.

In the course of this thesis, a new version of \curt will be developed, called
\acurt. This version is not only aware of its own state, but can also gather
information about the individuals it knows about. Furthermore, \acurt is
able to account for several possible readings of a certain sentence at once and
in parallel, which is a side-effect of the introduction of \term{intensionality}
into the system.

\acurt is able to parse and, according to the theoretical background of section
\ref{sec:theory}, correctly interpret sentences as the ones in
\ref{ex:whemba}-\ref{ex:whembc}.

\ex.
\a.  Mia knows Vincent snorts.\label{ex:whemba}
\b.  Mia knows who snorts.
\b.  Vincent believes that Mia likes Jody.\label{ex:whembc}

For this to work, the internal storage of \curt will have to be restructured
significantly in order to deal with more than one possible world or situation.
Its grammar will have to be enhanced so that it can handle \wh-embedding verbs
like `know' and `believe' and its inferencing mechanisms will have to be
enhanced.

The following section explains the problems at hand and existing solutions. A
subset thereof will be implemented in section \ref{sec:implementation}, which
will also raise some pragmatic issues discussed later on in section
\ref{sec:discussion}.

\section{Theory}\label{sec:theory}

@@

Following \cite{gs:q}, we will not use the natural English word
``question'' in its traditional sense, but distinguish between \term{questions},
\term{interrogative sentences} and \term{interrogative acts} in the following
way:

\begin{termin}
  An \term{interrogative} is a certain type of sentence in natural language
  distinguished by certain characteristics\footnote{To quote \cite{gs:q}: \quote{[An
  interrogative sentence is] characterized by word order, intonation, question
  mark, the occurrence of interrogative pronouns}.}
  whereas an \term{interrogative act} refers to the utterance of an
  \term{interrogative}. Finally, a
  \term{question} will denote the semantic entity uttered by the
  \term{interrogative act}, as it will be introduced in section
  \ref{sec:protoq}
\end{termin}

The following abbreviations will be used: 
\abbr{Ty2} stands for two-sorted type logic\footnote{See \cite{gallin:ty2} for a
discussion and \cite{gs:sqpa} for its usage in describing the semantics of
questions}. \abbr{IL} stands for intensional logic\footnote{As used by \cite{ptq}}.

%todo OUTLINE Presenting a conceptual overview, while arguing for and against stuff.

@@

@@

\prol programs can be self-modifying, or dynamic. In this way the program is
able to change its own state and thus store information.

@@

\curt parses sentences of natural language using an enhanced version of \pn{Cooper
storage}\footnote{See \cite{cooper:storage2} for the original implementation.} as
presented in \cite{keller:storage}. \cite{blackburnbos:cl1} call this version
\pn{Keller Storage}. That name will be used from here % FIXME sucks. Please change
to refer to this particular implementation as presented in
\cite{blackburnbos:cl1}.

\pn{Keller Storage} is already quite capable
of dealing with more complex constituents such as nested noun phrases, but it
has its limitations. In particular, it can only handle quantifier scope
ambiguities and does not take negation into account. Moreover, it overgenerates
with some quantifier scope ambiguities\footnote{\cite{blackburnbos:cl1} mention
sentences such as \ref{ex:kellersucks}, where Keller Storage produces six
instead of the desired five readings.
\ex. One criminal knows every owner of a hash bar\label{ex:kellersucks}

Clearly, \example{a hash bar} cannot take scope over \example{every owner}, but
Keller Storage predicts this reading.}.

In order to cope with more complicated quantifier configurations, where
constraints need to be enforced, and with other kinds of ambiguities,
Blackburn and Bos present an approach utilizing underspecified representations and
dominance relations, \pn{Hole Semantics}\footnote{Also presented in % todo:cite
%Bos' paper
}. However none of the cases discussed so far would require the enhanced
expressive power of \pn{Hole Semantics} it was decided to not change the \curt
implementation's current architecture which employs \pn{Keller Storage}.

@@

\section{Discussion}\label{sec:discussion}

We have seen, the Curt system is quite capable of treating questions this way.
Having implemented modality and possible worlds in such a fashion, let's go on
and implement some even cooler stuff (tense, aspect, negation, comparatives,
anaphora). Basing all this on a nice implementation of DRT (discussed in
\cite{kampreyle:drt}) like the one given in \cite{blackburnbos:cl2} might be
interesting. The Pronto \abbr{QA} system of \cite{prontoqa} already defines a
means of representing questions and queries against a Discourse Representation
Structure.
